other functionalities

### 2020-02-27 inst/gitbook/05_other-functionalities.Rmd
Added `char.diff` to the manual

>>>>

## `char.diff`

This is yet another function for calculating distance matrices.
There are many functions for calculating pairwise distance matrices in `R` (`stats::dist`, `vegan::vegdist`, `cluster::daisy` or `Claddis::MorphDistMatrix`) but this one is the `dispRity` one.
It is slightly different to the ones mentioned above (though not that dissimilar from `Claddis::MorphDistMatrix`) in the fact that it focuses on comparing discrete morphological characters and tries to solve all the problems linked to these kind of matrices (especially dealing with special tokens).

The function intakes a matrix with either `numeric` or `integer` (`NA` included) or matrices with `character` that are indeed `integers` (e.g.`"0"` and `"1"`).
It then uses a [bitwise operations](https://en.wikipedia.org/wiki/Bitwise_operations_in_C) architecture implemented in `C` that renders the function pretty fast and pretty modular.
This bitwise operations translates the character states into binary values.
This way, `0` becomes `1`, `1` becomes `2`, `2` becomes `4`, `3` becomes `8`, etc...
Specifically it can handle any rules specific to special tokens (i.e. symbols) for discrete morphological characters. For example, should you treat missing values `"?"` as `NA` (ignoring them) or as any possible character state (e.g. `c("0", "1")`?)? And how to treat characters with a ampersand (`"&"`)? `char.diff` can answer to all these questions!

Let's start by a basic binary matrix 4\*3 with random `integer`:

```{r}
## A random binary matrix
matrix_binary <- matrix(sample(c(0,1), 12, replace = TRUE), ncol = 4,
                        dimnames = list(letters[1:3], LETTERS[1:4]))
```

By default, `char.diff` measures the hamming distance between characters:

```{r}
## The hamming distance between characters
(differences <- char.diff(matrix_binary))
```

Note that the results is just a pairwise distance (dissimilarity) matrix with some special dual class `matrix` and `char.diff`.
This means it can easily be plotted via the disparity package:

```{r}
## Visualising the matrix
plot(differences)
```

You can check all the numerous plotting options in the `?plot.char.diff` manual (it won't be developed here).

The `char.diff` function has much more options however (see all of them in the `?char.diff` manual) for example to measure different differences (via `method`) or making the comparison work per row (for a distance matrix between the rows):

```{r}
## Euclidean distance between rows
char.diff(matrix_binary, by.col = FALSE, method = "euclidean")
```

We can however make it more interesting by playing with the different rules to play with different tokens.
First let's create a matrix with morphological characters as numeric `character`s:

```{r}
## A random character matrix
(matrix_character <- matrix(sample(c("0","1","2"), 30, replace = TRUE), ncol = 5,
                           dimnames = list(letters[1:6], LETTERS[1:5])))

## The hamming difference between columns
char.diff(matrix_character)
```

Here the characters are automatically converted into bitwise integers to be compared efficiently. We can now add some more special tokens like `"?"` or `"0/1"` for uncertainties between state `"0"` and `"1"` but not `"2"`:

```{r}
## Adding uncertain characters
matrix_character[sample(1:30, 8)] <- "0/1"

## Adding missing data
(matrix_character[sample(1:30, 5)] <- "?")

## The hamming difference between columns including the special characters
char.diff(matrix_character)
```

Note here that it detected the default behaviours for the special tokens `"?"` and `"/"`: `"?"` are treated as `NA` (not compared) and `"/"` are treated as both states (e.g. `"0/1"` is treated as `"0"` and as `"1"`).
We can specify both the special tokens and the special behaviours to consider via `special.tokens` and `special.behaviours`.
The `special.tokens` are `missing = "?"`, `inapplicable = "-"`, `uncertainty = "\"` and `polymorphism = "&"` meaning we don't have to modify them for now.
However, say we want to change the behaviour for `"?"` and treat them as all possible characters and treat `"/"` as only the character `"0"` (as an integer) we can specify them giving a behaviour function:

```{r}
## Specifying some special behaviours
my_special_behaviours <- list(missing = function(x,y) return(y),
                              uncertainty = function(x,y) return(as.integer(0)))

## Passing these special behaviours to the char.diff function
char.diff(matrix_character, special.behaviour = my_special_behaviours)
```

The results are quiet different as before! Note that you can also specify some really specific behaviours for any type of special token.

```{r}
## Adding weird tokens to the matrix
matrix_character[sample(1:30, 8)] <- "%"

## Specify the new token and the new behaviour
char.diff(matrix_character, special.tokens = c(weird_one = "%"),
                            special.behaviours = list(
                                weird_one = function(x,y) return(as.integer(42)))
                            )
```

Of course the results can be quiet surprising then... But that's the essence of the modularity. You can see more options in the function manual `?char.diff`!
